---
title: "Getting Started with EQuA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with EQuA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(EQuA)
```

# Status

This vignette demonstrates the pure-R reference implementation. The package is
still a development version, so numerical conventions are regression-tested
but the public interface may evolve during manuscript preparation.

# Required inputs

EQuA requires:

- chronological age;
- an existing scalar estimated age;
- optional numeric conditioning variables for Ex-EQuA.

The package does not require methylation beta matrices and does not calculate
the source age estimator.

# Simulated example data

```{r simulated-data}
dat <- EQuA::simulated_equa_data
str(dat)
```

The `sample_role` variable separates a calibration sample from a small
application sample.

```{r split-data}
calibration <- subset(dat, sample_role == "calibration")
application <- subset(dat, sample_role == "application")
```

# Fitting and prediction

```{r fit-and-predict}
fit <- fit_equa(
  chronological_age = calibration$chronological_age,
  estimated_age = calibration$estimated_age,
  method = "kernel_end",
  probs = seq(0.1, 0.9, by = 0.1)
)

profile <- predict(
  fit,
  new_estimated_age = application$estimated_age,
  chronological_age = application$chronological_age
)

summary(fit)
summary(profile)
head(as.data.frame(profile, format = "long"))
```

# Adding conditioning variables

Ex-EQuA uses kernel weighting to condition on estimated age and one or more
named numeric variables. Prediction covariates must have the same names and
order as the calibration covariates.

```{r ex-equa}
fit_ex <- fit_equa(
  chronological_age = calibration$chronological_age,
  estimated_age = calibration$estimated_age,
  covariates = calibration["z"],
  method = "kernel_ex",
  probs = c(0.25, 0.5, 0.75)
)

profile_ex <- predict(
  fit_ex,
  new_estimated_age = application$estimated_age,
  new_covariates = application["z"],
  chronological_age = application$chronological_age
)

head(as.data.frame(profile_ex))
```

# Public GEO example data

EQuA includes two compact real-data examples:

```{r public-geo-data}
gse40279 <- EQuA::gse40279_equa
gse50660 <- EQuA::gse50660_equa

dim(gse40279)
dim(gse50660)
head(gse40279)
```

Each object contains all public samples from its GEO series but retains only
the GSM accession, chronological age, and an upstream Horvath v1 estimated
age. The package does not contain the beta matrices and does not calculate the
clock.

For example, the first 500 GSE40279 samples can serve as a calibration set,
with the remaining samples treated as new observations:

```{r public-geo-fit}
geo_calibration <- gse40279[seq_len(500), ]
geo_application <- gse40279[-seq_len(500), ]

geo_fit <- fit_equa(
  chronological_age = geo_calibration$chronological_age,
  estimated_age = geo_calibration$estimated_age,
  method = "kernel_end",
  probs = c(0.1, 0.5, 0.9)
)

geo_profile <- predict(
  geo_fit,
  new_estimated_age = geo_application$estimated_age,
  chronological_age = geo_application$chronological_age
)

head(as.data.frame(geo_profile))
```

The split above is only a software demonstration, not a recommended scientific
design. For profiles of every observation in an analysis sample, use
`crossfit_equa()` so each observation is predicted strictly out of fold.
GSE50660 used 351 of 353 Horvath model CpGs and no cohort-mean or
K-nearest-neighbor imputation. See the installed help pages and the repository
data-provenance document before scientific use.

# Repeated cross-fitting

Cross-fitting returns profiles for the calibration observations using only
out-of-fold training data. Vignettes and automated checks use only a few
repeats so they finish quickly. Research analyses may request more repeats
explicitly.

```{r cross-fitting}
profile_cf <- crossfit_equa(
  chronological_age = calibration$chronological_age,
  estimated_age = calibration$estimated_age,
  method = "kernel_end",
  folds = 5,
  repeats = 3,
  seed = 2026,
  keep_repeats = FALSE,
  keep_fold_assignments = TRUE
)

summary(profile_cf)
head(profile_cf$fold_diagnostics)
```

# Diagnostics

Prediction output reports:

- marginal training-range violations;
- effective kernel sample size;
- nearest standardized calibration distance;
- maximum normalized kernel weight;
- low-support and weight-normalization flags;
- whether an explicitly requested fallback was used.

No nearest-neighbor fallback will occur implicitly.

```{r diagnostics}
head(equa_diagnostics(profile))
```

# Data privacy

An `equa_fit` object retains calibration-level EAD and conditioning values
needed for future kernel prediction. It does not retain subject identifiers,
and its print and summary methods do not display participant rows. The object
can nevertheless contain sensitive participant-level information; review
disclosure risk before saving or sharing it.

See `vignette("method-specification", package = "EQuA")` for the exact
statistical definitions, implementation provenance, and edge-case behavior.
